---
title: "R Overview for IT"
author: "Operations"
date: '`r format(Sys.Date())`'
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Overview

What is R?

R is a programming environment built for statistical (data science) applications.
Based on "S" and "Lisp"

- ["R"](https://www.r-project.org)
- ["S"](https://en.wikipedia.org/wiki/S_(programming_language))
- [Lisp](https://en.wikipedia.org/wiki/Lisp_(programming_language))

More
- [History of R](http://blog.revolutionanalytics.com/2017/10/updated-history-of-r.html)

## Getting R

- ["R"](https://www.r-project.org)
- [CRAN Mirrors](https://cran.r-project.org/mirrors.html)
- [CRAN Packages](https://cloud.r-project.org/web/packages/index.html)

## R - under the hood

- Base
- Stats
- Methods (for S3 and S4 objects)
- Graphics
- Linear Algebra 
    + LINPACK, LAPACK, others may be used

## R - Packages

- [CRAN Packages](https://cloud.r-project.org/web/packages/index.html)
- [CRAN Package Task Views](https://cloud.r-project.org/web/views/)
  
- Over 12,000 packages
    + Written by experts across many fields
    + Statistics, Operations Research, Data Science, Machine Learning, Data Mining


```{r load packages, echo = FALSE, include = FALSE}
library(rvest)
library(magrittr)
library(ggplot2)
library(plotly)
library(dplyr)
library(zoo)
```

## Setting up a plot
[Link to Code](https://rpubs.com/filipstachura/rpackages)
```{r CRAN growth data prep, echo = TRUE}
url <- "https://cran.r-project.org/web/packages/available_packages_by_date.html"

page <- read_html(url) %>%
  html_node("table") %>%
  html_table() %>%
  mutate(count = rev(1:nrow(.))) %>%
  mutate(Date = as.Date(Date)) %>%
  mutate(Month = format(Date, format="%Y-%m")) %>%
  group_by(Month) %>%
  summarise(published = min(count)) %>%
  mutate(Date = as.Date(as.yearmon(Month))) -> pkgs
```

## Example: Prepare plotly graph
```{r plotly plot, echo = TRUE}
margins = list(l = 100, r = 100, b = 100, t = 100, pad = 4)

plot_pltly <- pkgs %>%
  plot_ly(x = ~Date, y = ~published, name= "Published packages", 
          type="scatter", mode = "lines") %>%
  layout(title = "CRAN packages published over time.", 
         margin = margins) 
```

## Growth of CRAN Packages (plotly)
```{r CRAN growth plot plotly, echo = TRUE}
plot_pltly
```

## Example: Prepare ggplot graph
```{r ggplot plot, echo = TRUE}
plot_gg <- pkgs %>%
  ggplot(aes(x = Date, y = published)) +
  geom_point(color = 'dark cyan') +
  geom_line(color = 'red') +
  scale_x_date(date_breaks = "years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('CRAN packages published over time')
```

## Growth of CRAN Packages (ggplot)
```{r CRAN growth plot ggplot, echo = TRUE}
plot_gg
```

## More R Characteristics

- Object Oriented
- Lists, dataframes, matrices, vectors, strings (no scalars)
- Functions, modules, packages
- Functional programming
- Hooks into other languages 
    + for instance: C, C++, Fortran, Java, Perl, etc. 

## Basically ...

- Everything that exists in R is an object.
- Everthing that happens in R is a function call.
- Interfaces to other software are part of R.

    + [John Chambers](https://statweb.stanford.edu/~jmc4/vitae.html)

## MoRe

- Interpreted language 
    + can speed up with: 
        + more preprocessing in database
        + vectorized R operations 
        + compiled R  
        + C/C++ 
        
    + [Tips for faster R](https://robinsones.github.io/Making-R-Code-Faster-A-Case-Study/)

- Recent improvements in packages with common syntax ("Tidyverse")

## Working with R

- Development
    + R itself (R scripts)
    + IDE (R Studio, Eclipse, Visual Studio, others)
    + code version management with svn and git (github)

- Deployment
    + Use within IDE (local machine, server)
    + Batch script execution
    + Called from other platforms
    + Interactive application

- [Data Science Workflow and R](https://speakerdeck.com/jennybc/workflow-you-should-have-one)

## Getting data

- Excel, csv, tab delimited
- Web scraping
- SharePoint (XML, JSON)
- Various APIs (Google, Yahoo Finance, FRED, NOAA, etc etc)

- SQL options
    + MS Access (if we must)
    + SQL Server (connects externally; is also built into)
    + Oracle (connects externally; is also built into)
    + MySQL
    + Others: PostGres, NoSQL, MariaDB
    + etc. (pretty much any DB system where we can define a connection string)

## Support

- [r-project.org](https://www.r-project.org/help.html)
- [Stackoverflow](https://stackoverflow.com/questions/tagged/r)
- [R Community](http://fg2re.sellorm.com/community.html)
- [R Studio community](https://community.rstudio.com/)
- [R-bloggers Learning R](https://www.r-bloggers.com/how-to-learn-r-2/)
- Many social media groups (Twitter, Google+, others)
- Many, many sites devoted to using R (UCLA IDRE; cookbook-R, etc etc etc)

## Enterprise

- R Studio Enterprise
- Revolution R Enterprise (Microsoft)
- Similar - S Plus (Tibco Software)
- R within Microsoft SQL Server
- R within Oracle

## Making results available

- Report writing
    + can write to MS Word, PowerPoint 
    + can create pdf docs, ioslides, slidify, etc
    + can create text and rtf docs
    + markdown
    + notebooks (similar to Jupyter)

- Reading and writing from/to Excel, Googlesheets

- Can send e-mails, publish social media (Twitter)

- Web publishing and web apps

## Does R do graphs?

[R Graph Gallery](https://www.r-graph-gallery.com/all-graphs/)  

- Popular  
    + base R plotting (very powerful)
    + ggplot2 
        + based on Cleveland's 'Grammer of Graphics'
        + [some examples](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
        + [extensions to ggplot2](http://www.ggplot2-exts.org/gallery/)
    + [plotly](https://plot.ly/)
    + integration with javascript libraries (d3.js, raphael)

- Many, many others


## Data Mining and Machine Learning

- Methods (a few)
    + ANOVA, Linear Regression, Logistic Regression
    + Principle Components Analysis, Factor Analysis, Discriminant Analysis
    + Cluster Analysis, Neural Networks, Bayesian methods
    + Text Mining and Sentiment Analysis
    + Support Vector Machines, Random Forests, Gradient Boosting, Ensemble Methods
    + Time Series and Forecasting

- Machine Learning Integration
    + Keras, Tensorflow, MS Azure, Spark, Hadoop, options for parallel processing

## Web publishing options

- [OpenCPU](https://www.opencpu.org/)
    + HTML/CSS/Java/JavaScript

- [Shiny with Shiny Server](https://shiny.rstudio.com/) 
    + Translates (sort of) R code to HTML, CSS, Java, Javascript
    + Very popular, growing in use and capabilities
    + Provides static and dynamic/interactive reporting and app development
    + Shiny development is fast, flexible
    + Can run on any OS; currently using Linux  server at Essendant

## Web publishing examples

- [Shiny Gallery Examples](http://shiny.rstudio.com/gallery/)
    + [flexible web layout options](https://shiny.rstudio.com/articles/layout-guide.html)  
        + sidebar, fluid grid system, tabsets, nav lists, navbar pages 
    + [themes and bootstrap](http://rstudio.github.io/shinythemes/)
    + [Shiny Dashboard structure](https://rstudio.github.io/shinydashboard/)

## Is R popular?

- [IEEE Rankings 2017](https://spectrum.ieee.org/computing/software/the-2017-top-programming-languages)
[(Interactive)](https://spectrum.ieee.org/static/interactive-the-top-programming-languages-2017)

- [TIOBE Rankings 2017](https://www.tiobe.com/tiobe-index/)
    
- [Article on Data Science Software](http://r4stats.com/articles/popularity/)


## More Reading

- [Six Reasons to Use R for Business](https://www.r-bloggers.com/six-reasons-to-learn-r-for-business/)  

- [Field Guide to the R Ecosystem](http://fg2re.sellorm.com/)

## QUESTIONS ?

```{r 3d plot prep, echo = FALSE, include = FALSE}
library(rgl)
library(rglwidget)
library(knitr)
knit_hooks$set(webgl = hook_webgl)
```


## Rotatable 3D Plot
```{r 3d plot, r, webgl=TRUE}
next3d()
plot3d(mtcars$wt, mtcars$disp, mtcars$mpg, type="s", size=1, lit=TRUE, main = "Car Weight Vs Engine Displacement Vs Mileage",sub="3-D Plot")
```

## This presentation:

- Created with
    + R
    + rmarkdown (markdown package with inline R functionality)
        + ioslides format
        + uses knit ('knitr' package)
        + can also render to pdf (Beamer), MS Word, MS PowerPoint, slidy 
        + etc etc


- Published with
    + shiny
    + shiny-server
    